Analyzing TTC Delay Data to Improve Future Efficiency

### 2024-01-25 outputs/paper.qmd
Fix typo
Update paper per peer reviews

---
title: "Analyzing TTC Delay Data to Improve Future Efficiency"
abstract: "TTC delays have been increasing over time, and especially now in a period of overhaul in many parts of the system, it becomes important to analyze the effects of subway and bus delays in relation with each other as well as commuter loads. Using data published by the TTC and accessed through `opendatatoronto` we find that the greatest causes of delays are date and time, and equipment quality. Date and time factors are out of anyones control, but must be accounted for to improve future operational efficieny, while equipment quality has also been shown to be important, with mechanical failures heavily decreasing the reliability of buses."
thanks: "Code and data supporting this analysis is available at: https://github.com/prajogt/ttc_analysis.git"
author: "Timothius Prajogi"
date: today
date-format: long
format: pdf
editor: visual
toc: true
number-sections: true
bibliography: refrences.bib
---

@@

The main source public transportation of public transportation in the city of Toronto is the TTC (Toronto Transit Commission). The most common methods of this public transportation is subways and buses. For people who frequently use these methods, delays are common and expected [@CityNewsBGrade]. So much so that leaving early is leaving on time. With the delays getting worse and service becoming more inconsistent [@TorontoStarTTCDelays], it is important to analyze potential causes, which would therefore enable the TTC to explore potential solutions. One of such possible factors is the tight coupling of the subway and bus systems. When issues occur with the subway, the overflow is sent to buses, either from detoured commuter traffic, or resource reallocation in the form of converting regular route buses to shuttle buses [@TorontoObserverShuttleBuses]. Whether you, the reader, may be a commuter or not, having a good and reliable source of public transportation is crucial to reducing travel time for all city goers, not to mention its additional benefits for the environment.

@@

The datasets used for this report are the latest delay statistics from the 2023 year, `TTC SUbway Delay Data` [@citesubwaydata] and `TTC Bus Delay Data` [@citebusdata], both of which were published by the TTC.

Data was downloaded, cleaned, parsed, analyzed, and visualized using R [@citeR], a statistical programming language, with package support from `tidyverse` [@citetidyverse], a collection of libraries which included the following packages that were utilized:

@@

Both @tbl-bus_delay_data and @tbl-subway_delay_data contain columns "Date" and "Time" in which the delay had occurred, and an "Incident" in which the type of incident that had occurred out of these 4 general types: "Operator" (for driver/operator related issues), "Security/Safety" (for fire, police, trespassing, etc.), "Equipment/Mechanical" (for equipment, track, vehicle, etc., failures), and "Miscellaneous" (for diversions and other external issues).

For @tbl-bus_delay_data and @tbl-subway_delay_data the "Min Delay" and "Min Gap" are in minutes. Where "Min Delay" represents the delay in the schedule for the next bus and "Min Gap" is the scheduled time from the bus ahead of the following bus.

For @tbl-subway_delay_data the additional parameter "Line" represents the subway line in which the delay took place in. For those not familiar, these are the names of the different subway tracks the TTC utilizes.

In total @tbl-bus_delay_data contains 51,406 entries, whilst @tbl-subway_delay_data contains 22,465 entries, a significant difference.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vstime
#| fig-cap: "The amount of delays with respect to the time of day"
#| fig-subcap: ["Bus Delays", "Subway Delays"]
#| layout-ncol: 2

@@

bus_delay_data |>
  ggplot(aes(x = time)) +
  geom_histogram(fill = "blue", alpha = 0.75) +
  scale_x_time(breaks = scales::pretty_breaks(n = 12)) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

subway_delay_data |>
  ggplot(aes(x=time)) +
  geom_histogram(fill = "red", alpha = 0.75) +
  scale_x_time(breaks = scales::pretty_breaks(n = 12)) +
  theme_light() + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

@@

@fig-vsttime shows the difference in the trends of delays over time. Whilst bus delays heavily spike during rush hours where traffic alongside the buses has an impact on their operation, subway delays show a less significant increase in delays. Something important to note is the difference in scale that is present for all the graphs presented in this report, with the peak count of delays for @fig-vstime-2 being only around half of the peak count of delays for @fig-vstime-1.

@@

@fig-avg_num_vsday shows the similar trends that are present in the average number of delays that occur on certain days of the week. On average Sundays in 2023 produce the lowest amount of delays. Buses and subways average 101.00 and 50.75 delays respectively on Sundays. It seems that the added commuter load or the assumption of an increased commuter load causes more delays. We cannot say for certain which of the two it is, as it is equally likely that the reason that more delays occur on certain days is because more buses and trains are scheduled more frequently to accommodate the load of working commuters. Equally, it is possible that the increase in delays is caused by a human actor. As shown by @fig-vsincident, a large part of the delays for subways are security and safety issues which is almost entirely comprised of human effects. What we can say now however, is that more preparation and further analysis into commuter effects must be done by the TTC to lower the large amount of delays caused.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vsincident
#| fig-cap: "The average number of delays occurred due to incident types"
#| fig-subcap: ["Bus Delays", "Subway Delays"]
#| layout-ncol: 2

@@

bus_delays_by_incident <- read_csv(
   file = "data/summaries/total_num_bus_delays_by_incident.csv",
   show_col_types = FALSE
) 
bus_delays_by_incident |> 
  ggplot(aes(x=incident, y=n, fill=incident)) +
  geom_bar(stat = "identity") +
  xlab("Incident Type") +
  ylab("Delay Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

subway_delays_by_incident <- read_csv(
   file = "data/summaries/total_num_subway_delays_by_incident.csv",
   show_col_types = FALSE
) 
subway_delays_by_incident |> 
  ggplot(aes(x = incident, y = n, fill = incident)) +
  geom_bar(stat = "identity") +
  xlab("Incident Type") +
  ylab("Delay Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

@@

From @fig-vsincident-2 we can see that the leading cause of delay on the subways is "Security/Safety" incidents with 15,367 delays caused by such issues, with the next leading incident being "Equipment/Mechanical" issues with 4,005 delays occurred. "Security/Safety" covers track invasions, unsanitary trains, fire, etc. and "Equipment/Mechanical" covers any problems occurring with the functionality of the trains, the station, and the tracks. From personal experience, these results are expected. Comparatively, bus delays in @fig-vsincident-1 are majorly caused by "Equipment/Mechanical" issues with 18,786 caused. The second leading incident type is operator issues which covers any problems with the driver of the buses. These incidents account for 14,400 delays. These results pose an interesting connection.

From the TTC website, we know that they have been moving towards implementing automation in the operation of subway services since 2022[@TTCAutomation]. This has not been possible for buses which is why operator error is still high on the list. This could also be explained by the ratio of operators to commuters. Whilst each bus requires an operator, multiple cars on a subway only require 1-2 operators especially with the addition of automated signalling. Similarly, the lack of employee oversight within stations and subways can explain the difference in security and safety issues. Buses don't have much to oversee, however since stations are large and employees are only stationed within the entrance floor, prevention of disruptive behaviour is more commonplace as displayed by @fig-vsincident-2 above.

### By Line

@@

```
From @fig-vsline we see an expected result, where the longest (and oldest) lines with Yonge-University having 38 stations and Bloor-Danforth having 31 compared to Sheppard's 5 and Scarborough's 6 (previously existing) stations. With the load that these older lines are experiencing on a daily basis, the result from our analysis on delay causes is further supported. Constant maintenance and repair is required for these lines, and increased ridership also puts extra stress on the systems. Further engineering analysis is required to definitively know the extent of the effects of stress loads on these lines, and if those are the cause for the increase in delays.

## Conclusion

@@

## Reference

>>>>

---
title: "Analyzing TTC Delay Data to Improve Future Efficiency"
abstract: "TTC delays have been increasing over time, and especially now in a period of overhaul in many parts of the system, it becomes important to analyze the effects of subway and bus delays in relation with each other as well as commuter loads. Using data published by the TTC and accessed through `opendatatoronto` we find that the greatest causes of delays are date and time, and equipment quality. Date and time factors are out of anyones control, but must be accounted for to improve future operational efficieny, while equipment quality has also been shown to be important, with mechanical failures heavily decreasing the reliability of buses."
thanks: "Code and data supporting this analysis is available at: https://github.com/prajogt/ttc_analysis.git"
author: "Timothius Prajogi"
date: today
date-format: long
format: pdf
editor: visual
toc: true
number-sections: true
bibliography: references.bib
header-includes:
  \usepackage{placeins}
---

@@

The main source public transportation of public transportation in the city of Toronto is the TTC (Toronto Transit Commission). From this commission, the most common methods used are subways and buses. For people who frequently use these methods, delays are common and expected [@CityNewsBGrade], so much so that leaving early is leaving on time. With the delays getting worse and service becoming more inconsistent [@TorontoStarTTCDelays], it is important to analyze potential causes, which would therefore enable the TTC to explore potential solutions. One of such possible factors is the tight coupling of the subway and bus systems. When issues occur with the subway, the overflow is sent to buses, either from detoured commuter traffic, or resource reallocation in the form of converting regular route buses to shuttle buses [@TorontoObserverShuttleBuses]. Whether one uses these methods or not, having a good and reliable source of public transportation is crucial to reducing travel time for all city goers, not to mention its additional benefits for the environment.

@@

The datasets used for this report are the latest delay statistics from the 2023 year, `TTC Subway Delay Data` [@citesubwaydata] and `TTC Bus Delay Data` [@citebusdata], both of which were published by the TTC.

These delay statistics were downloaded, cleaned, parsed, analyzed, and visualized using R [@citeR], a statistical programming language, with package support from `tidyverse` [@citetidyverse], a collection of libraries which included the following packages that were utilized:

@@

Both the bus and subway delay dataset contain columns "Date" and "Time" in which the delay had occurred, and an "Incident" in which the type of incident that had occurred out of these 4 general types: "Operator" (for driver/operator related issues), "Security/Safety" (for fire, police, trespassing, etc.), "Equipment/Mechanical" (for equipment, track, vehicle, etc., failures), and "Miscellaneous" (for diversions and other external issues).

"Min Delay" and "Min Gap" are represented in minutes. Where "Min Delay" represents the delay in the schedule for the next bus and "Min Gap" is the scheduled time from the bus ahead of the following bus.

For the subway delay dataset, the additional feature "Line" represents the subway line in which the delay took place in. For those not familiar, these are the names of the different subway tracks the TTC utilizes. The 4 TTC lines operating in 2023 are labeled with their official name, however, the dataset also includes incidents that occurred at intersections between Yonge-University and Bloor-Danforth, which for the purposes of this report was considered separately. Additionally, lines which did not match any of the existing TTC lines were labeled "Other", as if they were significant enough to include in this dataset, they should have a relationship with the subway Incidents.

In total the bus delay dataset that @tbl-bus_delay_data samples contains 51,406 entries, whilst the subway delay dataset that @tbl-subway_delay_data samples contains 22,465 entries, a significant difference. To be precise, the amount of delays that have occurred in 2023 is 2.29 times the amount of delays that have occurred for the subway. If we further consider the filtered data, the difference amount of significant (in this case meaning with a delay time greater than 1 minute) is widened as will be discussed in @sec-issues_with_dataset.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vstime1
#| fig-cap: "The amount of delays with respect to the time of day"
#| fig-width: 4

@@

bus_delay_data |>
  ggplot(aes(x = time)) +
  geom_histogram(fill = "blue", alpha = 0.75) +
  scale_x_time(breaks = scales::pretty_breaks(n = 12)) +
  xlab("Time") +
  ylab("Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  theme(axis.text=element_text(size=12),
      axis.title=element_text(size=14,face="bold"))

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vstime2
#| fig-cap: "The amount of delays with respect to the time of day"
#| fig-width: 4

subway_delay_data |>
  ggplot(aes(x=time)) +
  geom_histogram(fill = "red", alpha = 0.75) +
  scale_x_time(breaks = scales::pretty_breaks(n = 12)) +
  xlab("Time") +
  ylab("Count") +
  theme_light() + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold"))

@fig-vstime1 and @fig-vstime2 shows the difference in the trends of delays over time. Whilst bus delays heavily spike during rush hours where traffic alongside the buses has an impact on their operation, subway delays show a less significant increase in delays. Subways do not have to deal with external traffic factors which is likely the cause of buses showing a larger amount of delays during those hours in relation with other times. Something important to note is the difference in scale that is present for all the graphs presented in this report, with the peak count of delays for @fig-vstime2 being only around half of the peak count of delays for @fig-vstime1.

@@

@fig-avg_num_vsday shows the similar trends that are present in the average number of delays that occur on certain days of the week. On average Sundays in 2023 produce the lowest amount of delays. Buses and subways average 101.00 and 50.75 delays respectively on Sundays. It seems that the added commuter load or the assumption of an increased commuter load causes more delays. We cannot say for certain which of the two it is, as it is equally likely that the reason that more delays occur on certain days is because more buses and trains are scheduled more frequently to accommodate the load of working commuters. Equally, it is possible that the increase in delays is caused by a human actor. As shown by @fig-vsincident1 and @fig-vsincident2, a large part of the delays for subways are security and safety issues which is almost entirely comprised of human effects. What we can say now however, is that more preparation and further analysis into commuter effects must be done by the TTC to lower the large amount of delays caused.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vsincident1
#| fig-cap: "The average number of delays occurred due to incident types"

@@

bus_delays_by_incident <- read_csv(
   file = "data/summaries/total_num_bus_delays_by_incident.csv",
   show_col_types = FALSE
) 
bus_delays_by_incident |> 
  ggplot(aes(x=incident, y=n, fill=incident)) +
  geom_bar(stat = "identity") +
  xlab("Incident Type") +
  ylab("Delay Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  guides(fill=guide_legend(title="Incident Type"))

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-vsincident2
#| fig-cap: "The average number of delays occurred due to incident types"

subway_delays_by_incident <- read_csv(
   file = "data/summaries/total_num_subway_delays_by_incident.csv",
   show_col_types = FALSE
) 
subway_delays_by_incident |> 
  ggplot(aes(x = incident, y = n, fill = incident)) +
  geom_bar(stat = "identity") +
  xlab("Incident Type") +
  ylab("Delay Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  guides(fill=guide_legend(title="Incident Type"))

From @fig-vsincident2 we can see that the leading cause of delay on the subways is "Security/Safety" incidents with 15,367 delays caused by such issues, with the next leading incident being "Equipment/Mechanical" issues with 4,005 delays occurred. "Security/Safety" covers track invasions, unsanitary trains, fire, etc. and "Equipment/Mechanical" covers any problems occurring with the functionality of the trains, the station, and the tracks. From personal experience, these results are expected. Comparatively, bus delays in @fig-vsincident1 are majorly caused by "Equipment/Mechanical" issues with 18,786 caused. The second leading incident type is operator issues which covers any problems with the driver of the buses. These incidents account for 14,400 delays. These results pose an interesting connection.

From the TTC website, we know that they have been moving towards implementing automation in the operation of subway services since 2022[@TTCAutomation]. This has not been possible for buses which is why operator error is still high on the list. This could also be explained by the ratio of operators to commuters. Whilst each bus requires an operator, multiple cars on a subway only require 1-2 operators especially with the addition of automated signalling. Similarly, the lack of employee oversight within stations and subways can explain the difference in security and safety issues. Buses don't have much to oversee, however since stations are large and employees are only stationed within the entrance floor, prevention of disruptive behaviour is more commonplace as displayed by @fig-vsincident2 above.

### By Line {#sec-by_line}

@@

From @fig-vsline we see an expected result, where the longest (and oldest) lines with Yonge-University having 38 stations and Bloor-Danforth having 31 compared to Sheppard's 5 and Scarborough's 6 (previously existing) stations. With the load that these older lines are experiencing on a daily basis, the result from our analysis on delay causes is further supported. Constant maintenance and repair is required for these lines, and increased ridership also puts extra stress on the systems. Further engineering analysis is required to definitively know the extent of the effects of stress loads on these lines, and if those are the cause for the increase in delays.

\FloatBarrier
## Conclusion

@@

\newpage
## References
